---
title: "Reading help in your language"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reading help in your language}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, comment = "#>")
```

This guide is for **R users** who would rather read documentation in their own
language. Nothing here changes how your code runs — only what `?function` shows.

## The whole thing in four lines

```{r}
remotes::install_github("adit-0132/rhelpi18n@dynamic-sexpr")  # 1. the engine
remotes::install_github("SOMEONE/gsocproposal.es")            # 2. a translation
# 3. RESTART R
Sys.setLanguage("es"); library(rhelpi18n); ?greet            # 4. read it
```

The rest explains each step.

## 1. Install the engine

**rhelpi18n** makes R look for a translation whenever you open a help page.
Install it once.

```{r}
install.packages("remotes")
remotes::install_github("adit-0132/rhelpi18n@dynamic-sexpr")
```

(Until merged upstream, use the `@dynamic-sexpr` branch; afterwards just
`remotes::install_github("eliocamp/rhelpi18n")`.)

## 2. Install the package and a translation

```{r}
install.packages("gsocproposal")                    # the package (however you normally would)
remotes::install_github("SOMEONE/gsocproposal.es")  # a Spanish translation module
```

Translations ship as their own small packages, usually named
`<package>.<language>`. Install as many as you like — one per language, or for
different packages; they don't conflict. If none exists for your package and
language, the original package help will be shown as normal.

## 3. Choose your language

Help language is controlled by the `LANGUAGE` environment variable. Base R's
`Sys.setLanguage()` is the convenient setter — it wraps `Sys.setenv(LANGUAGE =
...)` and returns the previous value:

```{r}
# this session only:
Sys.setLanguage("es")

# permanently: add a line to ~/.Renviron, then restart R
usethis::edit_r_environ()   # add:  LANGUAGE=es
```

On an R build without NLS support, or in a `C`/`POSIX` locale, `Sys.setLanguage()`
declines to set the variable — translated help still works there, so fall back to
`Sys.setenv(LANGUAGE = "es")`.

| Language | `LANGUAGE` |
|---|---|
| Spanish | `es` |
| French | `fr` |
| Brazilian Portuguese | `pt_BR` |
| Simplified Chinese | `zh_CN` |
| Traditional Chinese | `zh_TW` |
| Back to English | `en` |

Use **underscores, not hyphens** (`zh_CN`, not `zh-CN`), and be specific:
`zh_CN` and `zh_TW` are treated as different languages.

## 4. Restart R, then read

```{r}
# Session -> Restart R
library(rhelpi18n)
library(gsocproposal)
?greet
```

The help pane now shows the translated page, with any version/date-specific bits
filled in with the real, current value. Sections a translator hasn't done yet
stay in English — you always get correct information, never stale.

## Seeing the original English again

To see the help page in the original language, change your language to a language with no
translation installed (the default is `"en"`):

```{r}
Sys.setLanguage("en"); ?greet
```

## Switching languages

```{r}
Sys.setLanguage("fr"); ?greet   # French
```

No reinstall needed to switch — just change the variable (restart R if a page
seems stuck).

## Troubleshooting

- **Still English?** Did you *restart R* after installing? Check
  `Sys.getenv("LANGUAGE")` and that `library(rhelpi18n)` is loaded.
- **Some sections translated, some not?** That translation is incomplete, or the
  package was updated since it was written.
- **List what translations you have:**

```{r}
ip <- installed.packages(fields = c("Translates", "Language"))
ip[!is.na(ip[, "Translates"]), c("Package", "Translates", "Language")]
```

## How it works (short version)

When you open a help page, rhelpi18n notices the package and your `LANGUAGE`,
finds an installed translation module declaring that package + language, and
swaps the translated text in — re-inserting live values (dates, versions,
generated tables) from your actual installation. That last point is why a
translation stays correct after you reinstall or update the package.
