---
title: "Translating a package's help pages"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Translating a package's help pages}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, comment = "#>")
```

This guide is for **translators**: you want to translate the *documentation*
(help pages) of an R package into another language. You never touch the
package's code — you only edit text files.

The example package throughout is `gsocproposal`, a small demo that exercises
every rdlocal feature.

## 1. Get the package *source*

You need the package's **source** — a folder with a `DESCRIPTION` file and a
`man/` folder of `.Rd` help files — not the installed version.

```{r}
# clone from GitHub, or download+unzip the source tarball from CRAN
system("git clone https://github.com/someone/gsocproposal.git ~/translate/gsocproposal")
```

## 2. Generate the translation skeleton

Pick a language code (`es`, `fr`, `zh_CN`, `pt_BR`, …) and run:

```{r}
library(rdlocal)

i18n_module_create(
  package_path = "~/translate/gsocproposal",   # the SOURCE folder
  language     = "es",                          # your language code
  module_path  = "~/translate/gsocproposal.es"  # where to create the translation
)
```

You get a small module package:

```
gsocproposal.es/
  DESCRIPTION            # which package + language this translates
  R/translation.R        # machinery; don't touch
  inst/translations/     # ONE .yaml per help page -- your work goes here
  man_original/          # copies of the original .Rd, for reference
```

## 3. Anatomy of a translation file

Open `inst/translations/sexpr_install.yaml`:

```yaml
description:
  original: |
    Output below comes from an install-stage Sexpr (baked to text at install):
    {ISEXPR_0}
  translation: ~
  spans:
    '0': "installed under R 4.5.3"
```

| Field | What it is | Edit it? |
|---|---|---|
| `original` | the English text, as it appears in the help page | **no** |
| `translation` | `~` means empty — your translation goes here | **yes** |
| `spans` | example of what each `{ISEXPR_i}` held at generation time | **no** (a hint) |
| `ifdef` | on some pages; one entry per platform-specific branch | **yes** |

## 4. The one rule: keep the `{ISEXPR_…}` placeholders

Text that changes every install (a version, a date, a generated list) is
replaced by a placeholder like `{ISEXPR_0}`. **Copy each placeholder into your
translation unchanged** — move it where your language needs it, but never
translate, renumber, or delete it. The `spans:` entry tells you what it stands
for (an *example* — the real value differs per install).

```yaml
description:
  original: |
    Output below comes from an install-stage Sexpr (baked to text at install):
    {ISEXPR_0}
  translation: |
    La salida de abajo proviene de un Sexpr de etapa install:
    {ISEXPR_0}
  spans:
    '0': "installed under R 4.5.3"
```

### Platform-specific blocks (`ifdef`)

Pages with `#ifdef`/`#ifndef` show different text per platform. Translate each
numbered branch in the `ifdef:` table (and keep its `{ISEXPR_i}` in `original`):

```yaml
description:
  original: |
    Text below appears on everything except Windows:
    {ISEXPR_0}The rest always appears.
  translation: |
    El texto de abajo aparece en todo excepto Windows:
    {ISEXPR_0}El resto siempre aparece.
  ifdef:
    '0': ESTA LÍNEA SE MUESTRA EN UNIX Y MACOS (NO WINDOWS).
```

### Rare: literal braces

If the original shows `{{ISEXPR_0}}` with **double** braces, that's real text
the author wrote — keep the double braces as-is.

## 5. Test locally

```{r}
install.packages("~/translate/gsocproposal.es", repos = NULL, type = "source")
# RESTART R (Session -> Restart R), then:
library(rdlocal)
library(gsocproposal)
Sys.setLanguage("es")
?sexpr_install
```

Always **restart R after installing** — otherwise R may show a stale version.

## 6. Publish on GitHub

Your translation module is a small R package; put it on GitHub so others can
install it.

```{r}
system("cd ~/translate/gsocproposal.es && git init && git add . && git commit -m 'Spanish (es) translation'")
# then create the repo and push (e.g. with the gh CLI)
system("cd ~/translate/gsocproposal.es && gh repo create gsocproposal.es --public --source=. --push")
```

Tell users to install it and set `LANGUAGE=es` — see the companion vignette,
*Reading help in your language*.

## Keeping a translation current

When the package releases a new version, re-sync your module against the new
source:

```{r}
i18n_module_update(
  module_path  = "~/translate/gsocproposal.es",  # your module
  package_path = "~/translate/gsocproposal"       # the new source
)
```

Preview first with `dry_run = TRUE` — it reports what would change without
touching your files:

```
5 unchanged  |  2 fuzzy  |  1 new  |  1 dropped
```

- **unchanged** — kept exactly as you left them.
- **new** — a section added in the new version, waiting with an empty
  `translation:`.
- **dropped** — a section that no longer exists; its translation is removed.
- **fuzzy** — the original text changed. Your old translation is kept but marked
  `fuzzy: yes`, with the old text in `previous_original:` so you can see what
  changed. Update the `translation:`, then **delete the `fuzzy:` and
  `previous_original:` lines**. Until you do, readers see your translation with a
  small "may be out of date" note that reveals the current original.

## Language codes

Codes follow gettext's `ll_CC` form — a lowercase
[ISO 639 language code](https://www.gnu.org/software/gettext/manual/html_node/Usual-Language-Codes.html)
(or a [three-letter code](https://www.gnu.org/software/gettext/manual/html_node/Rare-Language-Codes.html)
for rarer languages), optionally plus an uppercase
[ISO 3166 territory code](https://www.gnu.org/software/gettext/manual/html_node/Country-Codes.html).
Use an **underscore**, not a hyphen (`zh_CN`, not `zh-CN`); `zh_CN` (Simplified)
and `zh_TW` (Traditional) are distinct. The module *name* gets dots automatically
(`gsocproposal.zh.CN`) — that's fine; the real tag lives in the module's
`Language:` field.
