---
title: "Translation of Rd"
author: "Elio"
date: "2024-02-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)


rd2char <- function(x, n = 1) {
  
  tag <- attr(x, "Rd_tag")
  
  if (is.null(tag)) {
    if (n == 1) {
      tag <- deparse(substitute(x))
    } else {
      tag <- "NULL"  
    }
    
  }
  if (!is.list(x)) {
    x <- gsub("\\n", "\\\\n", x)
    return(paste0("<", tag, "> ", '"', x, '"'))
  } 
  sep <- paste0("\n", strrep("  ", n), "└─")
  elements <- vapply(seq_along(x), function(i) paste0(sep, "[", i, "]", rd2char(x[[i]], n = n + 2)), FUN.VALUE = character(1))
  
  elements <- paste0(elements, collapse = "")  
  
  
  out <- paste0("<", tag, ">", elements)  
  
  return(out)
}


print.Rd <- function(x) {
  cat(rd2char(x, n = 1))
}

list2char <- function(x, title = NULL, n = 1) {
  
  if (is.null(title)) {
    if (n == 1) {
      title <- deparse(substitute(x))
    } else {
      title <- ""  
    }
    
  }
  if (!is.list(x)) {
    x <- gsub("\\n", "\\\\n", x)
    return(paste0("$", title, ": ", '"', x, '"'))
  } 
  sep <- paste0("\n", strrep("  ", n), "└─")
  elements <- vapply(seq_along(x), function(i) paste0(sep, "[", i, "]", list2char(x[[i]],
                                                                                  title = names(x)[i], 
                                                                                  n = n + 2)),
                     FUN.VALUE = character(1))
  
  elements <- paste0(elements, collapse = "")  
  
  out <- paste0("$", title, ": ", elements)  
  
  
  return(out)
}


print.rdlist <- function(x){
  title <- deparse(substitute(x))
  cat(list2char(x, title, n = 1))
}

```

After a lot of going around, the Rd file is converted either to HTML, text or pdf with `tools::Rd2HTML()`, `Rd2txt()` or `Rd2latex()`, respectively. 
These "take the output of `tools::parse_Rd()`, an Rd object, and produce a help page from it. As they are mainly intended for internal use, their interfaces are subject to change."

I think a good place to do the translation is right before this / during this. 
Potentially it could be better to use `Rd2HTML()` to do the translation because it would be able to include HTML-specific tricks like links, hover, warnings for outdated documentation, etc. 
But that function is extremely complex, so this would be a better first step. 

For example, the relevant snipped for HTML help, found on lines [555-557 in `tools:::httpd()`](https://github.com/r-devel/r-svn/blob/b2c5976be46a2470d2174cb9b04667771f3b08de/src/library/tools/R/dynamicHelp.R#L555-L557) is:

```r
Rd2HTML(utils:::.getHelpFile(file.path(path, helpdoc)),
out = outfile, package = dirpath,
dynamic = TRUE, outputEncoding = "UTF-8")
```

`utils:::.getHelpFile()` is not `tools::parse_Rd()`, but apparently the output is the same (`tools::parse_Rd()` might run at install time and then `utils:::.getHelpFile()` might fetch that result back). 

So, we could either sneak a `translate_rd()` function before `Rd2HTML()` or modify `Rd2HTML()` to do call `translate_rd()` first.
In any case, we need a function that takes an Rd object and some structure with the translated strings and returns a new Rd object with the strings replaced. 

```{r}
library(dummy)
help <- help("hello_world")

(rd <- utils:::.getHelpFile(help[1]))
```

This is just a list with some attributes and I implemented a custom printing method to make it easier to understand. 
Tags are printed between "<>" and elements that are lists have their contents printed nested under "└". 

Each element of the list is a section or tag that is another list with one element per line. 
The "Rd_tag" attribute holds the tag name. 

So, the first element:

```{r}
rd[[1]]
```


Is a list with Rd_tag `\\title`. 
The first (and only) element of the list is the title, which has the "Rd_tag" `TEXT`.
This is printed by my custom method like this: 

```{r}
print.Rd(rd[[1]])
```

The sixth element is the arguments section, which is more complex:

```{r}
print.Rd(rd[[6]])
```

Each argument is represented by an `\\item` tag, which has two elements, the name of the argument and then the description. 

Sections with multiple lines are split into different elements of the list, but the newline character is still preserved. 
Links to other functions are list elements with the tag `\\link`. 

Take a more complex help page:

```{r}
(rd <- utils:::.getHelpFile(help("mean")))
```

Pieces of text that are code are inside lists, as well as text with emphasis. 

How to arrange all this for translation is not trivial and now I'm feeling that the translation might be better done before parsing this complex structure. 
The translation strings should be something like 

```{r}
list(
  title = list(
    original = "Arithmetic Mean",
    translation = "Media aritmética"
  ),
  arguments = list(
    x = list(
      original = "an \\R object.  Currently there are methods for numeric/logical vectors and \\link[=Dates]{date}, \\link{date-time} and \\link{time interval} objects.  Complex vectors are allowed for \\code{trim = 0}, only.",
      translation = "un objecto \\R. Actualmente hay métodos para vectores lógicos/numéricos y de \\link[=Dates]{fecha}...."
    )
  )
)
```

That is, each section should be in its own place, with the original and the translation (the original is there in order to check that the translation is up-to-date), but the innards of each section shouldn't be parsed.

One solution would be to convert the rd object into something more manageable. 
For example:

```{r}
remove_newlines <- function(x) {
  x <- gsub("^\\n", "", x)
  x
}

## Creates text, including tags and options to each section. 
## Should be applied to:
## 1. each section
## 2. each argument name
## 3. each description
## 
to_text <- function(x) {
  tag <- attr(x, "Rd_tag")
  
  if (!is.null(tag) && !startsWith(tag, "\\")) {
    return(x[[1]])
  }

  inner <- vapply(x, to_text, FUN.VALUE = character(1))

  if (!is.null(tag)) {

    option <- attr(x, "Rd_option")
    if (!is.null(option)) {
      option <- paste0("[", option, "]") 
    }
    inner <- paste(inner, collapse = "}{")
    return(paste0(tag, option, "{", inner, "}"))
  }
  
  inner <- paste(inner, collapse = "")
  return(inner)
}

rd_tags <- function(help_db) {
  tags <- vapply(help_db, function(x) attr(x, "Rd_tag"), FUN.VALUE = character(1))
  gsub("\\\\", "", tags)
}

untranslatable <- c(
  "alias", 
  "name",
  "keyword",
  "concept",
  "usage"
)


make_text <- function(x) {
  tag <- attr(x, "Rd_tag")

  ## Here I treat the arguments section differently. 
  ## Maybe a better way would be to correctly parse \item{}{}
  ## elements
  if (tag != "\\arguments") {
    text <- vapply(x, to_text, FUN.VALUE = character(1))
    text <- paste(text, collapse = "") |> 
      remove_newlines() 
    if (tag %in% paste0("\\", untranslatable)) {
      return(text)
    }
    return(list(original = text, 
                translation = NULL))
  }
  
  text <- lapply(x, function(y) {
    tag <- attr(y, "Rd_tag")
    
    if (tag == "\\item") {
      description <- list(original = to_text(y[[2]]),
                          translation = NULL)
      attr(description, "name") <- to_text(y[[1]])
      return(description)
    }
    return(NULL)
  })
  
  text <- text[!vapply(text, is.null, FUN.VALUE = logical(1))]
  
  names <- vapply(text, function(x) attr(x, "name"), FUN.VALUE = character(1))
  names(text) <- names
  return(text)
}

rd_list <- function(rd) {
  list <- lapply(rd, make_text)
  names(list) <- rd_tags(rd)
  class(list) <- "rdlist"
  list
}


mean_rd_list <- rd_list(rd)

```


```{r}
mean_rd_list
```

There's some details here with whitespace and newlines that could be improved, but I think this is quite good!


```{r}
write_template <- function(rd_list) {

  rd_list <- rd_list[!(names(rd_list) %in% untranslatable)]
  
  yaml::write_yaml(rd_list, here::here("notes/mean.yaml"))  
}
write_template(mean_rd_list)
```

```{r}
mean_es <- yaml::read_yaml(here::here("notes/mean.es.yaml")) 
class(mean_es) <- "rdlist"
```


```{r}
translate <- function(original, translation) {
  
  translation <- translation[!(names(translation) %in% untranslatable)]
  
  sections <- names(original)
  sections <- sections[!(sections %in% untranslatable)]
  for (section in sections)  {
    
    
    if (!is.null(translation[[section]])) {
      if (is.character(original[[section]]$original)) {
        version_matches <- original[[section]]$original == translation[[section]]$original
        translation_exists <- !is.null(translation[[section]]$translation)
        
        if (version_matches && translation_exists) {
          original[[section]] <- translation[[section]]$translation
        } else {
          # If the translation is out of date?
          # For now, keep the original
          original[[section]] <- original[[section]]$original  
        }
      }
      
      if (is.list(original[[section]])) {
        original[[section]] <- translate(original[[section]], translation[[section]])
        
      }
    } else {
      
      original[[section]] <- original[[section]]$original  
    }
  }
  return(original)
}

(mean_translated <- translate(mean_rd_list, mean_es))

```

```{r}

section2char <- function(x) {
  
  if (is.character(x)) {
    return(x)
  } 
  
  paste(paste0("\\item{", names(x), "}{", unlist(x), "}"), 
        collapse = "\n") 
  
}
list2rdtext <- function(x) {
  texts <- vapply(x, section2char, FUN.VALUE = character(1))
  
  paste(paste0("\\", names(x), "{", texts, "}"), 
        collapse = "\n")
}

parse_rdlist <- function(rdlist) {
  text <- list2rdtext(rdlist)
  file <- tempfile()
  writeLines(text, file)
  tools::parse_Rd(file)
}


parse_rdlist(mean_translated)


```

```{r}
translate_helpfile <- function(helpfile) {
  
  original <- rd_list(helpfile)
  original$usage <- "mean(x, \\dots{})\n\n\\method{mean}{default}(x, trim = 0, na.rm = FALSE, \\dots{})\n"
  translation <- yaml::read_yaml("/home/elio/Documents/r-packages/internationalisation/rdlocal/notes/mean.es.yaml")
  
  translated <- translate(original, translation)
  
  parse_rdlist(translated)
}
```


```{r}
translate_helpfile(utils:::.getHelpFile(help("mean")))

```



